Associahedra for finite type cluster algebras \\ and minimal relations between $\b{g}$-vectors

### 2023-07-12 thirdPaper/draft/typeConeAssociahedra.tex
adresse Arnau
addresse Arnau
corrections type cone

\author{Arnau Padrol}
\address[Arnau Padrol]{Institut de Math\'{e}matiques de Jussieu - Paris Rive Gauche, Sorbonne Universit\'e, Paris}
\email{arnau.padrol@imj-prg.fr}
\urladdr{\url{https://webusers.imj-prg.fr/~arnau.padrol/}}

@@

\begin{abstract}
We show that the mesh mutations are the minimal relations among the $\b{g}$-vectors with respect to any initial seed in any finite type cluster algebra.
We then use this algebraic result to derive geometric properties of the $\b{g}$-vector fan: we show that the space of all its polytopal realizations is a simplicial cone, and we then observe that this property implies that all its realizations can be described as the intersection of a high dimensional positive orthant with well-chosen affine spaces.
This sheds a new light on and extends earlier results of N.~Arkani-Hamed, Y.~Bai, S.~He, and G.~Yan in type~$A$ and of V.~Bazier-Matte, G.~Douville, K.~Mousavand, H.~Thomas and E.~Y\i ld\i r\i m for acyclic initial seeds.

@@

This paper focuses on a surprising construction of the associahedron that recently appeared in mathematical physics.
Motivated by the prediction of the behavior of scattering particles, N.~Arkani-Hamed, Y.~Bai, S.~He, and G.~Yan recently described in~\cite[Sect.~3.2]{ArkaniHamedBaiHeYan} the kinematic associahedron.
It is a class of polytopal realizations of the classical associahedron obtained as sections of a high-dimensional positive orthant with well-chosen affine subspaces.
This construction provides a large degree of freedom in the choice of the parameters defining these affine subspaces, and actually produces all polytopes whose normal fan is affinely equivalent to that of J.-L.~Loday's associahedron~\cite{Loday} (see \cref{subsec:typeConeAsso}).
These realizations were then extended by V.~Bazier-Matte, G.~Douville, K.~Mousavand, H.~Thomas and E.~Y\i ld\i r\i m~\cite{BazierMatteDouvilleMousavandThomasYildirim} in the context of finite type cluster algebras using tools from representation theory of quivers.
More precisely, they fix a finite type cluster algebra~$\cal{A}$ and consider the real euclidean space~$\R^\cal{V}$ indexed by the set~$\cal{V}$ of cluster variables of~$\cal{A}$.
Starting from an acyclic initial seed~$\seed_\circ$, they consider a set~$\cal{M}$ of mesh mutations and compute the corresponding linear dependences between the $\b{g}$-vectors, which can be interpreted as linear spaces in~$\R^\cal{V}$.
Finally, they perturb these mesh linear spaces by a collection~$\b{\ell} \in \R_{>0}^{\cal{M}}$ of positive parameters and intersect the resulting perturbed mesh affine spaces with the positive orthant~$\R_{\ge 0}^\cal{V}$ (see \cref{subsec:typeConeCA}).
They show that the resulting polytope is always a generalized associahedron, and that its normal fan is the $\b{g}$-vector fan of~$\cal{A}$ with respect to the initial acyclic seed~$\seed_\circ$.
An implicit by-product of their construction is that the space of all polytopal realizations of the $\b{g}$-vector fan of~$\cal{A}$ with respect to the initial acyclic \mbox{seed~$\seed_\circ$ is a simplicial cone}.

\medskip
In this paper, we revisit, extend and explore further this construction using a reversed approach.
Given a complete simplicial fan~$\Fan$, we consider the space~$\typeCone(\Fan)$ of all its polytopal realizations.
This space was called type cone in~\cite{McMullen-typeCone} and deformation cone in~\cite{Postnikov, PostnikovReinerWilliams}, who studied the case when~$\Fan$ is the braid arrangement leading to the rich theory of generalized permutahedra (recently extended to all finite Coxeter arrangements in~\cite{ArdilaCastilloEurPostnikov}).
The type cone is known to be a polyhedral cone defined by a collection of inequalities corresponding to the linear dependences among the rays of~$\Fan$ contained in pairs of adjacent maximal cones of~$\Fan$ (see \cref{def:typeCone}).
Our approach is based on an elementary but powerful observation: for any fan~$\Fan$, all polytopal realizations of~$\Fan$ can be described as sections of a high dimensional positive orthant with a collection of affine subspaces parametrized by the type cone~$\typeCone(\Fan)$ (see \cref{prop:alternativePolytopalRealization}); if moreover the type cone~$\typeCone(\Fan)$ is a simplicial cone, it leads to a simple parametrization of all polytopal realizations of~$\Fan$ by a positive orthant corresponding to the facets of the type cone~$\typeCone(\Fan)$ (see \cref{coro:simplicialTypeCone}).
To prove that the type cone~$\typeCone(\Fan)$ is simplicial, we just need to identify which pairs of adjacent maximal cones of~$\Fan$ correspond to the facets of~$\typeCone(\Fan)$ and to show that the corresponding linear dependences among their rays positively span the linear dependence among the rays of any pair of adjacent maximal cones of~$\Fan$.
When applied to the $\b{g}$-vector fans of cluster algebras (see \cref{subsec:typeConeCA}), the type cone approach yields all polytopal realizations of the $\b{g}$-vector fans and thus efficiently revisits and extends the construction of~\cite{BazierMatteDouvilleMousavandThomasYildirim}.
This new perspective has several advantages, as our proof uniformly~applies~to:

\smallskip
\begin{itemize}
\item \textbf{any initial seed}, regardless of whether it is acyclic or not (see \cref{exm:typeConeCA,fig:clusterFans,fig:labelFacetDefiningInequalititiesCA1,fig:labelFacetDefiningInequalititiesCA2} for examples in types~$A_3$ and~$C_3$~cyclic). In contrast, the proof of~\cite{BazierMatteDouvilleMousavandThomasYildirim} only treats acyclic initial seeds although the approach has since been extended to any seed in~\cite{BazierMatte-thesis} (as previously announced by H.~Thomas in a lecture given at RIMS in June 2019).

\smallskip
\item \textbf{all finite type cluster algebras}, regardless of whether it is simply-laced or not (see \cref{exm:typeConeCA} for an example in type~$C_3$). In contrast, even with the acyclicity restriction, the result of~\cite{BazierMatteDouvilleMousavandThomasYildirim} is first proved in simply-laced cases ($A$, $D$ and~$E$) using representation theory of quivers. Extension to the non simply-laced cases can then be argued by a folding argument. This subtle and technical argument is presented in~\cite{ArkaniHamedHeLam}.

\smallskip
\item \textbf{any positive real-valued parameters}, regardless of whether they are rational or not. In contrast, the proof of~\cite{BazierMatteDouvilleMousavandThomasYildirim} naturally applies to rational parameters and then requires an approximation argument.
\end{itemize}

@@

Besides revisiting the construction of~\cite{ArkaniHamedBaiHeYan, BazierMatteDouvilleMousavandThomasYildirim} and extending it to any initial seed (acyclic or not) in any finite type cluster algebra (simply-laced or not), our type cone approach is also successful when applied to the $\b{g}$-vector fans of other families of generalizations of the associahedron.
In the present paper, we explore specifically \textbf{gentle associahedra}. These complexes where constructed in~\cite{PaluPilaudPlamondon-nonkissing} as polytopal realizations of finite non-kissing complexes of gentle bound quivers. They encompass two families of simplicial complexes studied independently in the literature: on the one hand the grid associahedra introduced by T.~K.~Petersen, P.~Pylyavskyy and D.~Speyer in~\cite{PetersenPylyavskyySpeyer} for a staircase shape, studied by F.~Santos, C.~Stump and V.~Welker~\cite{SantosStumpWelker} for rectangular shapes, and extended by T.~McConville in~\cite{McConville} for arbitrary grid shapes; and on the other hand the Stokes polytopes and accordion associahedra studied by Y.~Baryshnikov~\cite{Baryshnikov}, F.~Chapoton~\cite{Chapoton-quadrangulations}, A.~Garver and T.~McConville~\cite{GarverMcConville} and T.~Manneville and V.~Pilaud~\cite{MannevillePilaud-accordion}. The latter, sometimes called accordiohedra, have also~recently appeared in the mathematical physics litterature concerning scattering amplitudes~\cite{BanerjeeLaddhaRaman, Raman}, with realizations inspired by those in~\cite{ArkaniHamedBaiHeYan}. It was shown in~\cite{PaluPilaudPlamondon-nonkissing, BrustleDouvilleMousavandThomasYildirim} that the non-kissing complex of a gentle bound quiver provides a combinatorial model for the support $\tau$-tilting complex~\cite{AdachiIyamaReiten} of the associated gentle algebra. Even for finite non-kissing complexes, it turns out that the type cone of the non-kissing fan is not always simplicial (see \cref{exm:nonSimplicialTypeConeNKC}). We prove however that the type cone is always simplicial when the gentle bound quiver is brick (at least $2$ relations in any cycle) and $2$-acyclic (no oriented cycle of length~$2$). Moreover, we precisely describe the facet-defining inequalities of the type cone in terms of linear dependences of $\b{g}$-vectors involved in certain mesh relations. We thus automatically derive a construction similar to that of~\cite{ArkaniHamedBaiHeYan, BazierMatteDouvilleMousavandThomasYildirim} describing all polytopal realizations of the $\b{g}$-vector fans of these brick and $2$-acyclic gentle bound quivers as sections of a positive orthant by well-chosen affine spaces (see \cref{subsec:typeConeCA}). Again, this result mainly relies on finding which relations minimally generate the relations between the $\b{g}$-vectors in the non-kissing complex. We present both a purely combinatorial proof (see \cref{subsec:typeConeCA}) and a purely algebraic proof for (non-necessarily gentle) brick algebras making use of extriangulated categories (see \cref{sec:extricats}).

@@

\medskip
\enlargethispage{.5cm}
Further families of generalizations of the associahedron certainly deserve to be studied with our type cone approach.
We are investigating in particular graph associahedra\footnote{A preliminary version of this paper contained the full facet description of the type cones of graph associahedra. As they are not simplicial, they lie slightly beyond the scope of this paper, and will be studied elsewhere.}~\cite{CarrDevadoss, Postnikov, FeichtnerSturmfels, Zelevinsky}, brick polytopes~\cite{PilaudSantos-brickPolytope, PilaudStump-brickPolytope} and quotientopes~\cite{Reading-latticeCongruences, Reading-CambrianLattices, PilaudSantos-quotientopes}.
In fact, as already mentioned, it seems sometimes easier to find all polytopal realizations of a fan (by describing its type cone in terms of certain specific pairs of adjacent maximal cones) than to identify one specific explicit realization.
This is clearly the case when the type cone is simplicial as illustrated by the results of this paper: we naturally obtained all polytopal realizations of the $\b{g}$-vector fans of cluster algebras from cyclic seeds and of gentle algebras whose polytopality was only established very recently~\cite{HohlwegPilaudStella, PaluPilaudPlamondon-nonkissing}.
We believe that this approach gives reasonable hope that the polytopality of further fans could in the future be established using our type cone approach.
Tempting candidates are quotientopes for hyperplane arrangements beyond type~$A$, which are particularly interesting since the coarsenings of a fan essentially correspond to the faces of its type cone (see \cref{subsec:facesTypeCone}).

@@

\smallskip
\item The rays of the type cone of~$\Fan$ correspond to specific polytopes which form a positive Minkowski basis of all realizations of~$\Fan$ (see \cref{subsec:MinkowskiSums}). In particular, when the type cone is simplicial, any polytopal realization of~$\Fan$ has a unique representation (up to translation) as a Minkowski sum of positive dilates of these polytopes. For cluster algebras, it follows from~\cite[Sect.~6]{BazierMatteDouvilleMousavandThomasYildirim} that these polytopes are Newton polytopes of the $F$-polynomials. It would be interesting to obtain similar algebraic interpretations for the rays of the type cones of the non-kissing fans of gentle~algebras.

@@

\smallskip
\noindent
Another question related to this paper is to understand the complete realization space of the combinatorics of the associahedra.
The associahedra constructed in~\cite{ArkaniHamedBaiHeYan, BazierMatteDouvilleMousavandThomasYildirim} and the present paper describe all polytopal realizations of the $\b{g}$-vector fans, but there are other polytopal realizations of the combinatorial type of the associahedron with different normal fans.
Is it possible to similarly understand the complete realization space of the cluster complex, including all realizations not having the~$\b{g}$-vector fan as their normal fan?

@@

\noindent
Let~$\Fan$ be an essential complete polyhedral fan in~$\R^n$ (all definitions are recalled in~\cref{ssec:polytopes and fans}).
The type cone~$\typeCone(\Fan)$ of~$\Fan$ (\cref{ssec:type cone}), introduced by P. McMullen~\cite{McMullen-typeCone}, is a polyhedral cone that parametrizes the set of all possible polytopal realizations of~$\Fan$. These realizations can be described explicitly as affine sections of the non-negative orthant~$\R^N_{\geq0}$, where~$N$ is the number of rays of~$\Fan$ (\cref{prop:alternativePolytopalRealization}).  
If $\Fan$ is a simplicial fan and has the unique exchange relation property (\cref{def:uerp}), then its type cone is easier to describe and study.
And if moreover the type cone~$\typeCone(\Fan)$ is simplicial (\cref{rem:dimTypeCone}), then all possible polytopal realizations of~$\Fan$ can be described explicitely in terms of~$N-n$ positive real parameters as follows

@@

This gives a new point of view on~\cite{ArkaniHamedBaiHeYan,BazierMatteDouvilleMousavandThomasYildirim}, illustrated in \cref{subsec:typeConeAsso} for classical associahedra, that we then apply to two families of~$\b{g}$-vector fans generalizing the~$\b{g}$-vector fans of classical associahedra: the~$\b{g}$-vector fans of cluster algebras of finite type (generalized associahedra) in~\cref{subsec:typeConeCA}, and the~$\b{g}$-vector fans of gentle algebras (non-kissing associahedra, non-crossing associahedra, generalized accordiohedra) in~\cref{subsec:typeConeNKC}.

@@

And the second step follows from our main result in \cref{sec:clusterCategories}, which makes use of representation theory of finite dimensional algebras and additive categorification of cluster algebras.

@@

\begin{lemma}
\label{lem:MinkowskiSum}
For~$\b{h}, \b{h}' \! \in \R^N$, the polytope~$P_{\b{h} + \b{h'}}$ is the Minkowski sum of the polytopes~$P_\b{h}$~and~$P_{\b{h}'}$.
\end{lemma}

\begin{proof}
This follows from the definition of~$P_\b{h}$ and of Minkowski sums:
\begin{align*}
P_\b{h} + P_{\b{h}'} & = \set{\b{x} + \b{x}'}{\b{x} \in P_\b{h} \text{ and } \b{x}' \in P_{\b{h}'}} = \set{\b{x} + \b{x}'}{\b{G}\b{x} \le \b{h} \text{ and } \b{G}\b{x}' \le \b{h}'} \\
& = \set{\b{y} \in \R^n}{\b{G}\b{y} \le \b{h} + \b{h}'} = P_{\b{h} + \b{h}'}.
\qedhere
\end{align*}
\end{proof}

@@

In this section, we study the type cones of complete simplicial fans arising as normal fans of two families of generalizations of the associahedron: the generalized associahedra of finite type cluster algebras~\cite{FominZelevinsky-ClusterAlgebrasI, FominZelevinsky-ClusterAlgebrasII, FominZelevinsky-ClusterAlgebrasIV, HohlwegLangeThomas, HohlwegPilaudStella} and the gentle associahedra~\cite{PaluPilaudPlamondon-nonkissing}.
Both families contain the classical associahedra~$\Asso[n]$ constructed in~\cite{ShniderSternberg, Loday}.

@@

\pagebreak
We now focus on the normal fan of~$\Asso[n]$.
Since~$\Asso[n]$ lives in a hyperplane of~$\R^{n+1}$, its normal fan has a one-dimensional lineality (spanned by the all-ones vector).
To factor out the lineality, let~$\HH \eqdef \bigset{\b{x} \in \R^{n+1}}{\sum_{\ell \in [n+1]} \b{x}_\ell = 0}$ and ${\pi : \R^{n+1} \to \HH}$ denote the orthogonal projection.
We denote by~$\gvectorFan[n]$ the intersection of the normal fan of~$\Asso[n]$ with~$\HH$. 
Note that~$\gvectorFan[n]$ is an $n$-dimensional essential complete simplicial fan in~$\HH$, which is actually the normal fan of~$\pi(Asso[n])$ in~$\HH$. 
With the choice of a basis, we can identify $\HH$ with $\R^n$ and depict~$\gvectorFan[n]$ and~$\pi(\Asso[n])$ in~$\R^n$.

@@

Combining \cref{coro:simplicialTypeCone,coro:simplicialTypeConeCA,thm:extremalExchangeablePairsCA}, we derive the following description of all polytopal realizations of the cluster fan~$\gvectorFan[\B_\circ]$.
This result was stated in~\cite{BazierMatteDouvilleMousavandThomasYildirim} in the special situation of acyclic seeds in simply-laced types.
Note that our proof is quite different from~\cite{BazierMatteDouvilleMousavandThomasYildirim} as it relies on our type cone approach.

@@

We thank H.~Thomas for comments on a preliminary version of the paper and for informing us about the upcoming extended version of~\cite{BazierMatteDouvilleMousavandThomasYildirim}.
We are grateful to C.~Stump for reporting a typo in \cref{exm:typeConeCA}.
We also thank an anonymous referee for relevant comments.

>>>>

\author{Arnau Padrol}
\address[Arnau Padrol]{Departament de Matem\`atiques i Inform\`atica, Universitat de Barcelona}
\email{arnau.padrol@ub.edu}
\urladdr{\url{https://webusers.imj-prg.fr/~arnau.padrol/}}

@@

\begin{abstract}
We show that the mesh mutations are the minimal relations among the $\b{g}$-vectors with respect to any initial seed in any finite type cluster algebra.
We then use this algebraic result to derive geometric properties of the $\b{g}$-vector fan: we show that the space of all its polytopal realizations is a simplicial cone, and we then observe that this property implies that all its realizations can be described as the intersection of a high dimensional positive orthant with well-chosen affine spaces.
This sheds a new light on and extends earlier results of N.~Arkani-Hamed, Y.~Bai, S.~He, and G.~Yan in type~$A$ and of V.~Bazier-Matte, N.~Chapelet-Laguet, G.~Douville, K.~Mousavand, H.~Thomas and E.~Y\i ld\i r\i m for acyclic initial seeds.

@@

This paper focuses on a surprising construction of the associahedron that recently appeared in mathematical physics.
Motivated by the prediction of the behavior of scattering particles, N.~Arkani-Hamed, Y.~Bai, S.~He, and G.~Yan recently described in~\cite[Sect.~3.2]{ArkaniHamedBaiHeYan} the kinematic associahedron.
It is a class of polytopal realizations of the classical associahedron obtained as sections of a high-dimensional positive orthant with well-chosen affine subspaces.
This construction provides a large degree of freedom in the choice of the parameters defining these affine subspaces, and actually produces all polytopes whose normal fan is affinely equivalent to that of J.-L.~Loday's associahedron~\cite{Loday} (see \cref{subsec:typeConeAsso}).
These realizations were then extended by V.~Bazier-Matte, N.~Chapelet-Laguet, G.~Douville, K.~Mousavand, H.~Thomas and E.~Y\i ld\i r\i m~\cite{BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim} in the context of finite type cluster algebras using tools from representation theory of quivers.
More precisely, they fix a finite type cluster algebra~$\cal{A}$ and consider the real euclidean space~$\R^\cal{V}$ indexed by the set~$\cal{V}$ of cluster variables of~$\cal{A}$.
Starting from an acyclic initial seed~$\seed_\circ$, they consider a set~$\cal{M}$ of mesh mutations and compute the corresponding linear dependences between the $\b{g}$-vectors, which can be interpreted as linear spaces in~$\R^\cal{V}$.
Finally, they perturb these mesh linear spaces by a collection~$\b{\ell} \in \R_{>0}^{\cal{M}}$ of positive parameters and intersect the resulting perturbed mesh affine spaces with the positive orthant~$\R_{\ge 0}^\cal{V}$ (see \cref{subsec:typeConeCA}).
They show that the resulting polytope is always a generalized associahedron, and that its normal fan is the $\b{g}$-vector fan of~$\cal{A}$ with respect to the initial acyclic seed~$\seed_\circ$.
An implicit by-product of their construction is that the space of all polytopal realizations of the $\b{g}$-vector fan of~$\cal{A}$ with respect to the initial acyclic \mbox{seed~$\seed_\circ$ is a simplicial cone}.

\medskip
In this paper, we revisit, extend and explore further this construction using a reversed approach.
Given a complete simplicial fan~$\Fan$, we consider the space~$\typeCone(\Fan)$ of all its polytopal realizations.
This space was called type cone in~\cite{McMullen-typeCone} and deformation cone in~\cite{Postnikov, PostnikovReinerWilliams}, who studied the case when~$\Fan$ is the braid arrangement leading to the rich theory of generalized permutahedra (recently extended to all finite Coxeter arrangements in~\cite{ArdilaCastilloEurPostnikov}).
The type cone is known to be a polyhedral cone defined by a collection of inequalities corresponding to the linear dependences among the rays of~$\Fan$ contained in pairs of adjacent maximal cones of~$\Fan$ (see \cref{def:typeCone}).
Our approach is based on an elementary but powerful observation: for any fan~$\Fan$, all polytopal realizations of~$\Fan$ can be described as sections of a high dimensional positive orthant with a collection of affine subspaces parameterized by the type cone~$\typeCone(\Fan)$ (see \cref{prop:alternativePolytopalRealization}); if moreover the type cone~$\typeCone(\Fan)$ is a simplicial cone, it leads to a simple parameterization of all polytopal realizations of~$\Fan$ by a positive orthant corresponding to the facets of the type cone~$\typeCone(\Fan)$ (see \cref{coro:simplicialTypeCone}).
To prove that the type cone~$\typeCone(\Fan)$ is simplicial, we just need to identify which pairs of adjacent maximal cones of~$\Fan$ correspond to the facets of~$\typeCone(\Fan)$ and to show that the corresponding linear dependences among their rays positively span the linear dependence among the rays of any pair of adjacent maximal cones of~$\Fan$.
When applied to the $\b{g}$-vector fans of cluster algebras (see \cref{subsec:typeConeCA}), the type cone approach yields all polytopal realizations of the $\b{g}$-vector fans and thus efficiently revisits and extends the construction of~\cite{BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim}.
This new perspective has several advantages, as our proof uniformly~applies~to:

\smallskip
\begin{itemize}
\item \textbf{any initial seed}, regardless of whether it is acyclic or not (see \cref{exm:typeConeCA,fig:clusterFans,fig:labelFacetDefiningInequalititiesCA1,fig:labelFacetDefiningInequalititiesCA2} for examples in types~$A_3$ and~$C_3$~cyclic). In contrast, the proof of~\cite{BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim} only treats acyclic initial seeds although the approach has since been extended to any seed in~\cite{BazierMatte-thesis} (as previously announced by H.~Thomas in a lecture given at RIMS in June 2019).

\smallskip
\item \textbf{all finite type cluster algebras}, regardless of whether it is simply-laced or not (see \cref{exm:typeConeCA} for an example in type~$C_3$). In contrast, even with the acyclicity restriction, the result of~\cite{BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim} is first proved in simply-laced cases ($A$, $D$ and~$E$) using representation theory of quivers. Extension to the non simply-laced cases can then be argued by a folding argument. This subtle and technical argument is presented in~\cite{ArkaniHamedHeLam}.

\smallskip
\item \textbf{any positive real-valued parameters}, regardless of whether they are rational or not. In contrast, the proof of~\cite{BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim} naturally applies to rational parameters and then requires an approximation argument.
\end{itemize}

@@

Besides revisiting the construction of~\cite{ArkaniHamedBaiHeYan, BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim} and extending it to any initial seed (acyclic or not) in any finite type cluster algebra (simply-laced or not), our type cone approach is also successful when applied to the $\b{g}$-vector fans of other families of generalizations of the associahedron.
In the present paper, we explore specifically \textbf{gentle associahedra}. These complexes where constructed in~\cite{PaluPilaudPlamondon-nonkissing} as polytopal realizations of finite non-kissing complexes of gentle bound quivers. They encompass two families of simplicial complexes studied independently in the literature: on the one hand the grid associahedra introduced by T.~K.~Petersen, P.~Pylyavskyy and D.~Speyer in~\cite{PetersenPylyavskyySpeyer} for a staircase shape, studied by F.~Santos, C.~Stump and V.~Welker~\cite{SantosStumpWelker} for rectangular shapes, and extended by T.~McConville in~\cite{McConville} for arbitrary grid shapes; and on the other hand the Stokes polytopes and accordion associahedra studied by Y.~Baryshnikov~\cite{Baryshnikov}, F.~Chapoton~\cite{Chapoton-quadrangulations}, A.~Garver and T.~McConville~\cite{GarverMcConville} and T.~Manneville and V.~Pilaud~\cite{MannevillePilaud-accordion}. The latter, sometimes called accordiohedra, have also~recently appeared in the mathematical physics litterature concerning scattering amplitudes~\cite{BanerjeeLaddhaRaman, Raman}, with realizations inspired by those in~\cite{ArkaniHamedBaiHeYan}. It was shown in~\cite{PaluPilaudPlamondon-nonkissing, BrustleDouvilleMousavandThomasYildirim} that the non-kissing complex of a gentle bound quiver provides a combinatorial model for the support $\tau$-tilting complex~\cite{AdachiIyamaReiten} of the associated gentle algebra. Even for finite non-kissing complexes, it turns out that the type cone of the non-kissing fan is not always simplicial (see \cref{exm:nonSimplicialTypeConeNKC}). We prove however that the type cone is always simplicial when the gentle bound quiver is brick (at least $2$ relations in any cycle) and $2$-acyclic (no oriented cycle of length~$2$). Moreover, we precisely describe the facet-defining inequalities of the type cone in terms of linear dependences of $\b{g}$-vectors involved in certain mesh relations. We thus automatically derive a construction similar to that of~\cite{ArkaniHamedBaiHeYan, BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim} describing all polytopal realizations of the $\b{g}$-vector fans of these brick and $2$-acyclic gentle bound quivers as sections of a positive orthant by well-chosen affine spaces (see \cref{subsec:typeConeCA}). Again, this result mainly relies on finding which relations minimally generate the relations between the $\b{g}$-vectors in the non-kissing complex. We present both a purely combinatorial proof (see \cref{subsec:typeConeCA}) and a purely algebraic proof for (non-necessarily gentle) brick algebras making use of extriangulated categories (see \cref{sec:extricats}).

@@

\medskip
\enlargethispage{.5cm}
Further families of generalizations of the associahedron certainly deserve to be studied with our type cone approach.
We are investigating in particular graph associahedra\footnote{A preliminary version of this paper contained the full facet description of the type cones of graph associahedra. As they are not simplicial, they lie slightly beyond the scope of this paper. These results now appeared in a paper of A.~Padrol, V.~Pilaud and G.~Poullot on type cones of graph associahedra and nestohedra~\cite{PadrolPilaudPoullot-deformedNestohedra}.} and nestohedra~\cite{CarrDevadoss, Postnikov, FeichtnerSturmfels, Zelevinsky}, brick polytopes~\cite{PilaudSantos-brickPolytope, PilaudStump-brickPolytope} and quotientopes~\cite{Reading-latticeCongruences, Reading-CambrianLattices, PilaudSantos-quotientopes}.
In fact, as already mentioned, it seems sometimes easier to find all polytopal realizations of a fan (by describing its type cone in terms of certain specific pairs of adjacent maximal cones) than to identify one specific explicit realization.
This is clearly the case when the type cone is simplicial as illustrated by the results of this paper: we naturally obtained all polytopal realizations of the $\b{g}$-vector fans of cluster algebras from cyclic seeds and of gentle algebras whose polytopality was only established very recently~\cite{HohlwegPilaudStella, PaluPilaudPlamondon-nonkissing}.
We believe that this approach gives reasonable hope that the polytopality of further fans could in the future be established using our type cone approach.
Tempting candidates are quotientopes for hyperplane arrangements beyond type~$A$, which are particularly interesting since the coarsenings of a fan essentially correspond to the faces of its type cone (see \cref{subsec:facesTypeCone}).

@@

\smallskip
\item The rays of the type cone of~$\Fan$ correspond to specific polytopes which form a positive Minkowski basis of all realizations of~$\Fan$ (see \cref{subsec:MinkowskiSums}). In particular, when the type cone is simplicial, any polytopal realization of~$\Fan$ has a unique representation (up to translation) as a Minkowski sum of positive dilates of these polytopes. For cluster algebras, it follows from~\cite[Sect.~6]{BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim} that these polytopes are Newton polytopes of the $F$-polynomials. It would be interesting to obtain similar algebraic interpretations for the rays of the type cones of the non-kissing fans of gentle~algebras.

@@

\smallskip
\noindent
Another question related to this paper is to understand the complete realization space of the combinatorics of the associahedra.
The associahedra constructed in~\cite{ArkaniHamedBaiHeYan, BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim} and the present paper describe all polytopal realizations of the $\b{g}$-vector fans, but there are other polytopal realizations of the combinatorial type of the associahedron with different normal fans.
Is it possible to similarly understand the complete realization space of the cluster complex, including all realizations not having the~$\b{g}$-vector fan as their normal fan?

@@

\noindent
Let~$\Fan$ be an essential complete polyhedral fan in~$\R^n$ (all definitions are recalled in~\cref{ssec:polytopes and fans}).
The type cone~$\typeCone(\Fan)$ of~$\Fan$ (\cref{ssec:type cone}), introduced by P. McMullen~\cite{McMullen-typeCone}, is a polyhedral cone that parameterizes the set of all possible polytopal realizations of~$\Fan$. These realizations can be described explicitly as affine sections of the non-negative orthant~$\R^N_{\geq0}$, where~$N$ is the number of rays of~$\Fan$ (\cref{prop:alternativePolytopalRealization}).  
If $\Fan$ is a simplicial fan and has the unique exchange relation property (\cref{def:uerp}), then its type cone is easier to describe and study.
And if moreover the type cone~$\typeCone(\Fan)$ is simplicial (\cref{rem:dimTypeCone}), then all possible polytopal realizations of~$\Fan$ can be described explicitely in terms of~$N-n$ positive real parameters as follows

@@

This gives a new point of view on~\cite{ArkaniHamedBaiHeYan,BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim}, illustrated in \cref{subsec:typeConeAsso} for classical associahedra, that we then apply to two families of~$\b{g}$-vector fans generalizing the~$\b{g}$-vector fans of classical associahedra: the~$\b{g}$-vector fans of cluster algebras of finite type (generalized associahedra) in~\cref{subsec:typeConeCA}, and the~$\b{g}$-vector fans of gentle algebras (non-kissing associahedra, non-crossing associahedra, generalized accordiohedra) in~\cref{subsec:typeConeNKC}.

@@

The second step follows from our main result in \cref{sec:clusterCategories}, which makes use of representation theory of finite dimensional algebras and additive categorification of cluster algebras.

@@

\begin{lemma}
\label{lem:MinkowskiSum}
For any~$\b{h}, \b{h}' \in \typeCone(\Fan)$, the polytope~$P_{\b{h} + \b{h'}}$ is the Minkowski sum of the polytopes~$P_\b{h}$~and~$P_{\b{h}'}$.
\end{lemma}

\begin{proof}
As~$\b{h}, \b{h}' \in \typeCone(\Fan)$, the fan~$\Fan$ is the normal fan of both~$P_\b{h}$ and~$P_{\b{h}'}$, and thus of their Minkowski sum~$P_\b{h} + P_{\b{h}'}$.
Hence~$P_\b{h} + P_{\b{h}'} = P_{\b{h}''}$ for some~$\b{h}'' \in \typeCone(\Fan)$.
Moreover, we have~${\b{h}'' = \b{h} + \b{h}'}$ since for any ray~$\ray$ of~$\Fan$,
\begin{align*}
\b{h}''_{\ray} 
& = \max\set{\dotprod{\ray}{\b{x}''}}{\b{x}'' \in P_{\b{h}''}} = \max\set{\dotprod{\ray}{\b{x}''}}{\b{x}'' \in P_{\b{h}} + P_{\b{h}'}} \\
& = \max\set{\dotprod{\ray}{\b{x}}}{\b{x} \in P_{\b{h}}} + \max\set{\dotprod{\ray}{\b{x}'}}{\b{x}' \in P_{\b{h}'}} = \b{h}_{\ray} + \b{h}'_{\ray}.
\qedhere
\end{align*}
\end{proof}

@@

\enlargethispage{.6cm}
In this section, we study the type cones of complete simplicial fans arising as normal fans of two families of generalizations of the associahedron: the generalized associahedra of finite type cluster algebras~\cite{FominZelevinsky-ClusterAlgebrasI, FominZelevinsky-ClusterAlgebrasII, FominZelevinsky-ClusterAlgebrasIV, HohlwegLangeThomas, HohlwegPilaudStella} and the gentle associahedra~\cite{PaluPilaudPlamondon-nonkissing}.
Both families contain the classical associahedra~$\Asso[n]$ constructed in~\cite{ShniderSternberg, Loday}.

@@

We now focus on the normal fan of~$\Asso[n]$.
Since~$\Asso[n]$ lives in a hyperplane of~$\R^{n+1}$, its normal fan has a one-dimensional lineality (spanned by the all-ones vector).
To factor out the lineality, let~$\HH \eqdef \bigset{\b{x} \in \R^{n+1}}{\sum_{\ell \in [n+1]} \b{x}_\ell = 0}$ and ${\pi : \R^{n+1} \to \HH}$ denote the orthogonal projection.
We denote by~$\gvectorFan[n]$ the intersection of the normal fan of~$\Asso[n]$ with~$\HH$. 
Note that~$\gvectorFan[n]$ is an $n$-dimensional essential complete simplicial fan in~$\HH$, which is actually the normal fan of~$\pi(\Asso[n])$ in~$\HH$. 
With the choice of a basis, we can identify $\HH$ with $\R^n$ and depict~$\gvectorFan[n]$ and~$\pi(\Asso[n])$ in~$\R^n$.

@@

Combining \cref{coro:simplicialTypeCone,coro:simplicialTypeConeCA,thm:extremalExchangeablePairsCA}, we derive the following description of all polytopal realizations of the cluster fan~$\gvectorFan[\B_\circ]$.
This result was stated in~\cite{BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim} in the special situation of acyclic seeds in simply-laced types.
Note that our proof is quite different from~\cite{BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim} as it relies on our type cone approach.

@@

We thank H.~Thomas for comments on a preliminary version of the paper and for informing us about the upcoming extended version of~\cite{BazierMatteChapelierLaguetDouvilleMousavandThomasYildirim}.
We are grateful to C.~Stump for reporting a typo in \cref{exm:typeConeCA}.
We also thank an anonymous referee for relevant comments.
